---
title: "Mini Data-Analysis Deliverable 1"
output: html_document
---

# Welcome to my (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1. Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

2. Load the packages below.

```{r, message = FALSE, warning = FALSE}
library(datateachr)
library(tidyverse)
```

3. Make a repository in the https://github.com/stat545ubc-2021 Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Vincenzo as collaborators. 

4. When you go to submit, submit a URL to your repository to canvas. 

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

+ *apt_buildings*: Acquired courtesy of The City of Toronto’s Open Data Portal. It currently has 3455 rows and 37 columns.

+ *building_permits*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 20680 rows and 14 columns.

+ *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

+ *flow_sample*: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.

+ *parking_meters*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.

+ *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

+ *vancouver_trees*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

+ We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

+ Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

1: *apt_buildings*    
2: *cancer_sample*    
3: *flow_sample*    
4: *steam_games*    

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*. 

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

```{r}
### Exploring dataset 1: apt_buildings ### 
class(apt_buildings)
glimpse(apt_buildings)
```

**Attributes for the *apt_buildings* include:**

- Class of the data set: `c("tbl_df", "tbl", "data.frame")`, which means it is a tibble that has been coerced from a dataframe. 
- Number of rows: 3455
- Number of columns: 37
- Types of data in columns: 9 numeric types, 28 character types.
- Columns names
- Contents of variables in each column

**Comments:**

- This data set consists of different building information of 3455 apartments, including the year they were built and registered, number of units, facility conditions and so on. 
- Some columns that are `character` type only consist of 2 factors as "YES" or "NO". 
- It's possible that we can use it to study the relationship between the age of the buildings and the characteristics.

```{r}
### Exploring dataset 2: cancer_sample ### 
class(apt_buildings)
glimpse(cancer_sample)
```

**Attributes for the *cancer_sample* include:**

- Class of the data set: `c("tbl_df", "tbl", "data.frame")`, which means it is a tibble that has been coerced from a dataframe. 
- Number of rows: 569
- Number of columns: 32
- Types of data in columns: 31 numeric types, 1 character types (`diagnosis`).
- Columns names
- Contents of variables in each column

**Comments:**

- This data set includes 569 observations of diagnostic cancer data. 
- Besides the ID of samples and the ultimate diagnosis of the breast mass ("M" = malignant, "B" = benign), the columns are mostly **quantitative features** derived from the sample image. 
- Both statistic and clinical knowledge might be required to understand the variables and decide the ways to analyze and interpret the data.

```{r}
### Exploring dataset 3: flow_sample ### 
class(apt_buildings)
glimpse(flow_sample)
```
**Attributes for the *cancer_sample* include:**

- Class of the data set: `c("tbl_df", "tbl", "data.frame")`, which means it is a tibble that has been coerced from a dataframe. 
- Number of rows: 218
- Number of columns: 7
- Types of data in columns: 4 numeric types, 3 character types.
- Columns names
- Contents of variables in each column

**Comments:**

- This data set includes 218 observations of flow rate extrema data and additional information regarding the date and conditions of data collection. 
- If we see the `extreme_type` and the flow rate `flow` as dependent variable and the others as possible factors, we can have multiple angles to study the relationship between them by correlations. 


```{r}
### Exploring dataset 4: steam_games ### 
class(steam_games)
glimpse(steam_games)
```

**Attributes for the *steam_games* include:**

- Class of the data set: `c("spec_tbl_df", tbl_df", "tbl", "data.frame")`, as a tibble subclass.
- Number of rows: 40833
- Number of columns: 21
- Types of data in columns: 4 numeric types, 17 character types.
- Columns names
- Contents of variables in each column

**Comments:**

- This `spec_tbl_df` subclass is new and special to me. According to a documantation, it "differs from a regular tibble only in that the `spec` attribute (which holds the column specification) is lost as soon as the object is subset and a normal `tbl_df` object is returned." 
- It has 40833 observations, which is the most out of the 4 data sets that I've chosen.
- The columns of this set contains free text information as the url, names, developer and other details of the games. There's also common features across them that can be summarized, such as `types` and `genre`. There are also numeric features like `achivements` and `price` that seem interesting.  

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation. 

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1. Plot the distribution of a numeric variable.
2. Create a new variable based on other variables in your data (only if it makes sense)
3. Investigate how many missing values there are per variable. Can you find a way to plot this?
4. Explore the relationship between 2 variables in a plot.
5. Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6. Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7. Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8. Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->


### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
